---
title: "Intensive Intervention FTIC Characteristics"
author: "Jay Kim"
date: "3/24/2022"
output: html_document
---

The following FTIC tier groups are historically less well-prepared, and show poorer performance (or drop out) in their first year after entering UWF. This document summarizes their characteristics and looks for any patterns among  tier 4.



```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```
 
## Overall Tier Scores

**Tier 1**: score above 92 (old 90)

**Tier 2**: score above 87 (old 84)

**Tier 3**: score above 80 (old 77)

**Tier 4: score above 72 (old 69)**

**Tier 5**: score below 72 (old 69)
 
$$Tier score = (higher test score*0.03125) + (12.5*hsgpa capped at 4)$$
 
### Trends in Tier and Cohort 
 
```{r echo=FALSE}
library(readr)
library(tidyverse)
FTIC_ENROLLMENT_FIN_GROUPED2 <- read_csv("G:/Shared drives/HMCSE-PAM Lab/Jay's Space/_DataShaping/FTIC_ENROLLMENT_FIN_GROUPED2.csv")

#prop.table(table(FT_tier_sum$FTIC_APRCT, FT_tier_sum$FTIC_DEMO_TIME_FRAME),2)
FT_tier <- FTIC_ENROLLMENT_FIN_GROUPED2 %>%  filter(UWF_YearCode =="UWFYear1") %>% 
     mutate(APPLICANT_TIER = ifelse(is.na(APPLICANT_TIER), "NA", APPLICANT_TIER)) %>% 
    group_by(FTIC_UWFID) %>% 
    filter(FTIC_DEMO_TIME_FRAME == min(FTIC_DEMO_TIME_FRAME)) 
addmargins(table(COHORT=FT_tier$FTIC_Cohort, TIER=FT_tier$APPLICANT_TIER))
prop.table(table(COHORT=FT_tier$FTIC_Cohort, TIER=FT_tier$APPLICANT_TIER),1)
 
# which(duplicated(FT_tier$FTIC_UWFID))
#    
# app2017to2022_all_load_12_2021_V0 <- read_csv("G:/Shared drives/HMCSE-PAM Lab/Jay's Space/2021 Active Projects/08/ACCEPTANCE RATE/app2017to2022.all.load.12.2021_V0.csv")  %>%
#     group_by(UWFID) %>% filter(APP_NO == max(APP_NO) & (APP_DT == max(APP_DT))) %>% filter(!duplicated(UWFID))
# app_info <- app2017to2022_all_load_12_2021_V0
# which(duplicated(app_info$UWFID))
# 
# FT_tier_app <- merge(FT_tier, app_info, by.x="FTIC_UWFID", by.y="UWFID", all.x = T) %>% select(-contains("PROB"))
# which(duplicated(FT_tier_app$FTIC_UWFID))
# glimpse(FT_tier_app)
# test <- FT_tier_app[FT_tier_app$FTIC_Cohort=="2017","APP_DT"]
# max(test, na.rm = T)
# write.csv(FT_tier_app, "FT_tier_app.csv", row.names = FALSE)
```

```{r include=FALSE}
library(readr)
library(tidyverse)
mytierdata <- read_csv("FT_tier_app.csv") %>%
    mutate(APP_DT = as.Date(APP_DT, "%m/%d/%Y"), DEC_DATE= as.Date(DEC_DATE, "%m/%d/%Y")) %>% 
    mutate(APP_DT_month = format(as.Date(APP_DT), "%Y/%m")) %>% 
    mutate(DOB1 =  as.Date(DOB, "%m/%d/%Y"), FTIC_Age = round(as.numeric(difftime(APP_DT, DOB1, units = "weeks"))/52.25,0),
           DaysDecision= round(as.numeric(difftime(DEC_DATE, APP_DT, units = "days")) ,0)) %>% 
    mutate(GRADHSDATE =  as.Date(HS_GRAD_DT, "%m/%d/%Y"), GradHSYear = round(as.numeric(difftime(APP_DT, GRADHSDATE, units = "weeks"))/52.25,1)) %>% mutate_at(c(30:37), replace_na, 0) %>% 
    mutate(GradHSYear = ifelse(GradHSYear <= 0, 0, GradHSYear)) %>% 
    mutate(PELL_VERF = as.character(PELL_VERF)) %>% 
    mutate_if(is.character, as.factor) 
#Hmisc::describe(mytierdata)
aidtypenewcols <- mytierdata %>% select(30:37) %>% mutate_at(c(1:8), ~ifelse(.x == 0, "No","Yes")) 
names(aidtypenewcols) <- paste("Is.", names(aidtypenewcols), sep="")

filtered_data <- cbind(mytierdata,aidtypenewcols) %>%     
    mutate_if(is.character, as.factor) 
#addmargins(table(filtered_data$Is.GrantTypes)) 
#colnames(filtered_data)
```

## Data Description and Terms

**Terms**: 

 * APRCT: Above/below 2.00 in current (1st) term

 * GradHSYear: time (in months) from HS graduation to UWF application (each month is 0.1)

 * Days To Decision: time (in days) from aUWF application to decision (each day is 1.0)

**Financial Aid Type**

- Aid Categories: grants, institutional, private, other state, Bright Future, mixed

 * Grant: ARPG, CRRS, PELL, etc. 

 * Bright Future: State funded (Florida merit scholars funds)

 * Institutional: CARS, ACEX, ARG, FAOS, etc. (institutionally funded)

 * Private: PGSD#. (privately funded)

 * Other State: HELI, THEO, etc. (state funded but not bright future)

 * Loan: FDSL,FDUL, etc.

 * Mixed: state, federal, or third-party sponsor funded

**Aid Ratio**: proportion of aid type based on total aid amount (i.e. if "0.3 grant" then 30% of aid came from the grant category)

**Risk-FTIC**: Below 2.00 in first term


## Descriptive Statistics


* **Cohort**: from 2017 to 2021

* **Tier**: 4 

* **Demographic Information**: ethnicity, gender, county, high school name, and state 

* Performance/college information from their 1st term at UWF

### Summary Table by Cohort

```{r choose tier,   echo=FALSE}
filtered_data1 <- filtered_data %>%  filter(APPLICANT_TIER == 4) # filtered tier level
#addmargins(table(filtered_data1$Is.GrantTypes)) 
library(gtsummary)

factor_tier <- filtered_data1 %>% mutate(ST_group = ifelse(ST =="FL", "FL",ifelse(ST=="AL","AL","Others"))) %>% 
    select(FTIC_Cohort,FTIC_Age,GradHSYear,DaysDecision,FTIC_PriorHours, ST_group, is.factor,  -contains("DT"),-contains("DATE") ,-HS_CEEB, -ZIP,-DOB, -PROGRAM_CODE, -HS_NAME,-HS_CNTY,   -PROGRAM_DESC,-CURRICULUM_COLL,-CURRICULUM_DEPT, -UWF_YearCode,- APP_PAID,-APP_STATUS,-HS_CNTY, -CNTY,-RACE,-NATION, -GENDER, -APP_DT_month, contains("Is."),GrantTypes,LoanTypes,TotalFinAid_Term,-FTIC_County,-ST, -DEC_CODE, -ADMIT_TYPE, -OFFER_CODE,
           contains("scholar"))  
#colSums(is.na(factor_tier))
#addmargins(table(factor_tier$Is.GrantTypes, factor_tier$FTIC_Cohort)) 

#glimpse(factor_tier)
sumtab <- factor_tier %>%  mutate(FTIC_Cohort = as.factor(FTIC_Cohort))  %>% 
    tbl_summary( by = FTIC_Cohort,
                statistic = all_continuous() ~ "{mean} ({sd}) {min} {max}",
                missing = "no"
                            ) %>% add_n()
sumtab
#glimpse(factor_tier)
```

### Application Date by Cohort

```{r}
library(plotly)
library(lubridate)


appdate <- filtered_data1 %>%  
    ggplot() + geom_bar(aes(x=APP_DT_month, fill=factor(FTIC_Cohort))) + 
    coord_flip() + 
    facet_grid(~FTIC_Cohort)

ggplotly(appdate)


```
* Students tended to apply later in 2021 - the bulk applied in January.

* By contrast, from 2017 to 2020, most applications came in October and November. 


### Days from Application to Decision by Cohort and Applicatioin Number

```{r}
#hist(filtered_data1$DaysDecision)
decisiondays <- filtered_data1 %>% #filter(DaysDecision <= 200) %>% 
    ggplot() + geom_histogram(aes(x=DaysDecision ,fill=factor(APP_NO))) + coord_flip() +facet_grid(~FTIC_Cohort)
ggplotly(decisiondays)
```
* Most decisions were made within 100 days.

* There are some applicants with multiple applications which, combined, within 25 days for decision.


### Applied Before/After graduating high school by Some Financial Aid (Yes/No)

```{r}

gradhs <- filtered_data1 %>%  
    mutate(AfterGradHS=ifelse(GradHSYear !=0, "AppliedAfterGradHS","AppliedBeforeGradHS")) %>%  
    group_by(AfterGradHS, Is.TotalFinAid_Term) %>% dplyr::summarise(Count=n()) %>%  arrange(-Count) %>%
    ggplot(aes(x=reorder( AfterGradHS,Is.TotalFinAid_Term), y= Count, fill=Is.TotalFinAid_Term)) + geom_col() +
    scale_fill_manual(values = c("orange","gray"))
#summary(table(gradhs$FTIC_APRCT, gradhs$AfterGradHS))
ggplotly(gradhs)  
# library(doBy)
# summaryBy(c( "AGE") ~ AfterGradHS, data = gradhs)
```
* Most students without financial aid were traditional applicants (they applied before graduating HS).


### Summary of GPA and Prior Hours by Hs Name and Ethnicity

```{r}

race.prior <- filtered_data1 %>%  
    mutate(FTIC_PriorHours= ifelse(is.na(FTIC_PriorHours), 0, FTIC_PriorHours)) %>%            group_by(HS_NAME,FTIC_Ethnicity) %>% 
    dplyr::summarise(Count=n(),MIN=min(FTIC_PriorHours), MeanPriorHrs= round(mean(FTIC_PriorHours),2),
                     SD= round(sd(FTIC_PriorHours),2),
                     MAX=max(FTIC_PriorHours), MeanAge=round(mean(FTIC_Age, na.rm=T),2),
                     MeanGPA=round(mean(FTIC_UWFGPA),2)) %>% 
    arrange(-Count) %>%  filter(Count >= 5) 

knitr::kable(race.prior)
```


### Entry College and Gender

```{r}

college <- filtered_data1 %>% group_by(FTIC_College, FTIC_Gender) %>% summarise(Count=n()) %>% 
ggplot(aes(x= reorder(FTIC_College, FTIC_Gender) ,y=Count, fill=FTIC_Gender )) + geom_bar(stat = "identity") +
    coord_flip() 

ggplotly(college)

```
* HMCSE and COB Retained the most male FTIC.

* UKCOH retained the most female FTIC.



### Entry Department (cutoff >= 10)

```{r}

department <- filtered_data1 %>% group_by(FTIC_Department) %>% summarise(Count=n()) %>% filter(Count>=10) %>% 
ggplot(aes(x= reorder(FTIC_Department, Count) ,y=Count, fill=FTIC_Department )) + 
    geom_bar(stat = "identity") +
    coord_flip()

ggplotly(department)
```
* The most popular departments (in order) for Tier 4 by enrollment are biology, Business Administration, Nursing, Undeclared, and Computer Science.



### Entry Program by 1st Term UWF GPA (APRCT) (Above 2.00)

```{r}

program <- filtered_data1 %>% group_by(FTIC_ProgramCIPDesc, FTIC_APRCT) %>% summarise(Count=n()) %>% arrange(FTIC_ProgramCIPDesc,FTIC_APRCT,-Count) %>% 
    filter(Count >= 10) %>% 
ggplot(aes(x= reorder(FTIC_ProgramCIPDesc, Count) ,y=Count, fill=FTIC_APRCT )) + 
    geom_bar(stat = "identity") +
    coord_flip()

ggplotly(program)
```
* The largest groups of at-risk FTIC Below 2.00 came from the Business/Commerce general, Nursing, Biomedical Science and Cyber security programs.



### Top High school (cutoff >= 4) & Entry Program

```{r}
hs <- filtered_data1 %>% group_by(FTIC_ProgramCIPDesc, HS_NAME) %>% 
    mutate(FTIC_ProgramCIPDesc = substring(FTIC_ProgramCIPDesc,0,10)) %>% 
    dplyr::summarise(Count=n()) %>% arrange(-Count) %>% 
    filter(Count >= 3) %>% 
    ggplot() + geom_col(aes(x=reorder(FTIC_ProgramCIPDesc,HS_NAME), y= Count, fill=HS_NAME)) +#+coord_flip()
    #theme_bw(base_size = 9)+
    theme(axis.text = element_text(angle = 20))
ggplotly(hs)


```

* Top 3 most popular programs by high school are Nursing, Business/Commerce, General, and Cyber Security.

* Most students from NA HS name (GED) and Fort Walton HS entered the Nursing program.

* Most Gulf Breeze HS and JM Tate HS graduates entered the Business/Commerce General program.

* Most West FL HS graduates entered the Cyber security program.

### Residence Hall and some type of aid

```{r}

resident <- filtered_data1 %>% group_by(FTIC_ResidenceHall, Is.TotalFinAid_Term) %>% summarise(Count=n()) %>% 
     ggplot(aes(x= reorder(FTIC_ResidenceHall, Is.TotalFinAid_Term) ,y=Count, fill=Is.TotalFinAid_Term )) + 
    geom_bar(stat = "identity") +
    coord_flip() + scale_fill_manual(values = c("darkgray","lightblue"))

ggplotly(resident)

```
* Most students who did not have any financial aid were commuters or stayed in Martin Hall


### Tri-County by Grant Type Aid

```{r}
hscnty <- filtered_data1 %>% group_by(FTIC_County, Is.GrantTypes) %>% 
    dplyr::summarise(Count=n()) %>% arrange( -Count ) %>% 
    filter(Count >= 5) %>% 
    ggplot(aes(x=reorder(FTIC_County,Count), y= Count, fill=Is.GrantTypes)) + geom_col( ) +
    coord_flip()
ggplotly(hscnty)

```

* Most were came from tri-county and Non-FL with grants.



### High School Name and Ethnicity

```{r}

hsrace <- filtered_data1 %>% group_by( HS_NAME,FTIC_Ethnicity ) %>% 
    dplyr::summarise(Count=n()) %>% 
    arrange( -Count) %>% 
    filter(Count >= 5) %>% 
ggplot() + geom_col(aes(x= HS_NAME, y=Count, fill=FTIC_Ethnicity)) +coord_flip()
ggplotly(hsrace)

```

* Most African Americans came from not reporting a high school name (GED), WF HS Tech, and Milton HS.

* Most Hispanics came from GED, Navarre HS, and Fort Walton Beach HS.

* Most whites and "two or more ethnicity" came from JM Tate HS.


### Total Financial Aid Amount & APRCT

```{r}
aid.race <- ggplot(filtered_data1,aes(x=TotalFinAid_Term, fill=FTIC_APRCT)) +
    geom_histogram() #+ scale_fill_gradient(low="black", high="blue")
ggplotly(aid.race)
```

* A little less than 30% of FTIC did not have any type of aid, and were the bulk of the below 2.00 group and course withdrawn.

* Among students with aid, the count of at-risk for each group is fairly consistent.



### Financial Information: Total Financial Aid

 $$AidRatio = Aid Type Amount/Total Aid Amount$$

```{r}
library(plotly)
aid_ratio <- filtered_data1 %>% mutate_at(c(30:36), funs(./TotalFinAid_Term)) %>% 
    mutate_at(c(30:36), funs(round(., digits = 2))) %>% mutate_at(c(30:36), replace_na, 0) %>% 
    select(30:36) %>% pivot_longer(everything(), names_to = "AidType", values_to = "Ratio") %>%
    filter(Ratio > 0) %>%  # filtered no aid type
    ggplot(aes(x=Ratio,   fill=factor(AidType)) ) + geom_histogram() +labs(title = "Aid Type Ratio")

 ggplotly(aid_ratio)

```

* For those with only one type of aid, most were grants (152), loans (80), and UWF funded (50).

* Very few FTIC have other types of financial aid.

* Among those with more than one aid type, grants contributed more than 50% and loans contributed between 25 to 45% of their total financial aid.

### Financial Information by Group: Grant & Loan Types and APRCT

```{r}
library(plotly)
alltype <- filtered_data1[,c(18,32:33,37)] %>% tidyr::pivot_longer(c(2,3,4), names_to = "AidType", values_to = "Amount") %>%
    filter(!is.na(Amount) ) %>% 
    #mutate(Amount = round(Amount,2)) %>% 
  ggplot( aes(x= Amount, fill=FTIC_APRCT , colour=FTIC_APRCT)) + geom_histogram() + #facet_grid(~AidType) 
    #coord_flip() + 
    facet_grid(~AidType)
 ggplotly(alltype)
#colnames(filtered_data1)
```

* Most FTIC's aid came from grants and loans.

* Most at-risk FTIC came from the group with no aid.

### The Aid type for the **at-risk** FTIC (Below 2.00)


```{r}
aid.risk <- filtered_data1 %>% 
    filter(FTIC_UWFGPA < 2.00) %>% 
    select(84:90) %>%
    pivot_longer(c(1:7), names_to = "AidType", values_to = "Aid_Yes_No") %>% filter(!is.na(Aid_Yes_No)) %>% 
    ggplot(aes(x=AidType, fill=Aid_Yes_No)) + geom_bar(position="dodge") + facet_grid(~AidType) +
    labs(title = "Financial Aid Type among the risk-FTIC" ) + xlab(label ="") + coord_flip()
 
ggplotly(aid.risk) 
 
```

* Most aid for the below 2.00 group came from grants and loans.

 

```{r eval=FALSE, include=FALSE}
granttype <- read_csv("G:/Shared drives/HMCSE-PAM Lab/Jay's Space/2022 Active Projects/01/ACEX_CAL/granttype.csv")
ids <- filtered_data1 %>% select("UNIV_ROW_ID"=FTIC_UWFID)
granttypeID <-  merge(ids, granttype, by="UNIV_ROW_ID", all.x = T)
addmargins(table(granttypeID$DETAIL_CODE, granttypeID$REPT_TIME_FRAME))
```



### Pivot Table (APRCT by Ethnicity and Gender (%))

* The highest proportion of the below 2.0 group came from African American males (41%) with some type of grant.

```{r}

library(rpivotTable)
rpivotTable(filtered_data1, aggregatorName = "Count as Fraction of Rows",
            rows = c("Is.TotalFinAid_Term", "FTIC_Gender","FTIC_Ethnicity"),
             cols = c("FTIC_APRCT"),
            height = "500px", overflow="scroll",
             subtotals = T, rendererName = "Table With Subtotal")
```


